What We Eval in the Shadows

### 2021-08-13 paper/main.tex
Section 7: symbolic field reading and writing
Fixes

The R language was intended to be extensible. The combination of lazy
evaluation, \k{substitute} and \k{eval} are some of the tools given to
developers to this end. To understand \emph{why} developers use \eval, we
manually inspected top 117 sites which contribute to 90\% of the calls in the
corpus. In this section we show 10 real-world examples from this batch. The
goal is to illustrate the variety of use cases that employ \eval in the R
ecosystem. The code snippets are simplified for clarity.

\paragraph{Extended Interfaces} Sometimes it is convenient to allow users to
extend the behavior of a library function. \Eval allows clients to provide a
simple text string that shall be executed in the context of the function. The
example here allows users to provide formulas as strings, such as
\k{"exp(-x^2/2)"}, the contract being that the text can refer to some variable
\k{x} that will be set up by the called function.\footnote{The function is used
to generates a sequence of random numbers using the adaptive rejection sampling
algorithm and the string formula is used to describe the target density.} In
this case users could provide an anonymous function.

@@

This combination of \k{match.call} and \eval to pass most of the call to
another function is also used to created wrappers. For example, the
\k{write.csv} uses it to forward most of the parameters to the more general
\k{write.table} function, only setting the CSV-related parameters:

@@

The most probable reason for this is to ease maintenance of such
wrappers.\footnote{This was speculated by the Brian Ripley, the author of the
function, in a private conversation.} Wrapping functions in R is somewhat
harder than in other languages because of the default arguments, the varargs
and is missing arguments.

% TODO: update to explain why the eval is needed or remove
\paragraph{Formula Resolution}
Many packages use \eval to evaluate formulas. The example extracts the
\k{"formula"} attribute attached to the \k{pdMat} object and evaluates it.

\begin{minipage}{.95\textwidth}
  \medskip\underline{\c{nlme::formula.pdMat}:}
\begin{lstlisting}
 function(x, asList, ...) eval(attr(x, "formula"))
\end{lstlisting}\medskip
\end{minipage}

@@

The \k{glue} package contains a variant that leverages the non-standard scoping
as well allowing to resolve interpolation in an alternative environment. For
example, the following snippet pipes the subsetted data frame into the
\k{glue_data} function that uses it to resolve the symbols in string
interpolation.

@@

The cases where \eval appears to be overkill include simple expression shapes
such as \k{X}, \k{V}, \k{\$}, \k{<-}, and \k{FUN}. Variable lookup in a remote
environment can be performed by the builtin \k{get} function, a similar case
can be made for assignment and the \k{assign} function. There is a dedicated
\k{do.call} function to perform function calls. However, the unnecessary uses of
\eval often can only be uncovered by manual analysis.

@@

Another example, from \k{coin::copyslots}, constructs an expression that access
fields of an object and copy them. This can be replaced by symbolic field reads
and writes.

@@

% TODO: remove - does not bring anything new or interesting
As a last example, \k{configr::config.funs.par} constructs three expressions
that are then fed to \eval. Semantically equivalent code would use \k{get}
to lookup \k{fun}, and then directly execute the three expressions.

\begin{lstlisting}
 config.funs.par <- function(fun="",...) {
   args.all <- as.list(match.call())
   args.all <- args.all[names(args.all) != ""]
   parameters <- list()
   t1 <- sprintf("parameters <- names(as.list(args(%s)))", fun)
   t2 <- "parameters <- parameters[parameters != '']"
   t3 <- "parameters <- args.all[names(args.all) %in% parameters]"
   for (j in c(t1,t2,t3))
    eval(parse(text = j))
   return(parameters)
 }
\end{lstlisting}

The challenge for automation is that such examples are not easy to find. Dynamic
analysis is limited due to coverage issues, and static analysis of R remains an
open problem.

>>>>

The R language was intended to be extensible. The combination of lazy
evaluation, \k{substitute} and \k{eval} are some of the tools given to
developers to this end. To understand \emph{why} developers use \eval, we
manually inspected top 117 sites which contribute to 90\% of the calls in the
corpus. In this section we show 9 real-world examples from this batch. The
goal is to illustrate the variety of use cases that employ \eval in the R
ecosystem. The code snippets are simplified for clarity.

\paragraph{Behavioral Extension} Sometimes it is convenient to allow users to
extend the behavior of a library function. \Eval allows clients to provide a
simple text string that shall be executed in the context of the function. The
example here allows users to provide formulas as strings, such as
\k{"exp(-x^2/2)"}, the contract being that the text can refer to some variable
\k{x} that will be set up by the called function.\footnote{The function is used
to generates a sequence of random numbers using the adaptive rejection sampling
algorithm and the string formula is used to describe the target density.} In
this case users could provide an anonymous function instead.

@@

This combination of \k{match.call} and \eval for passing most of the current
call's argument to another function is also used to created wrappers. For
example, the \k{write.csv} uses it to forward most of the parameters to the
more general \k{write.table} function, only setting the CSV-related
parameters:

@@

The most probable reason for this is to ease maintenance of such
wrappers.\footnote{This was speculated by the Brian Ripley, the author of the
function, in a private conversation.} Wrapping functions in R is somewhat
harder than in other languages because of the default argument values, the
varargs and the missing arguments.

% % TODO: update to explain why the eval is needed or remove
% \paragraph{Formula Resolution}
% Many packages use \eval to evaluate formulas. The example extracts the
% \k{"formula"} attribute attached to the \k{pdMat} object and evaluates it.

% \begin{minipage}{.95\textwidth}
%   \medskip\underline{\c{nlme::formula.pdMat}:}
% \begin{lstlisting}
%  function(x, asList, ...) eval(attr(x, "formula"))
% \end{lstlisting}\medskip
% \end{minipage}

@@

The \k{glue} package contains a variant that leverages the non-standard scoping
allowing to resolve interpolation in an alternative environment. For example,
the following snippet pipes the subsetted data frame into the \k{glue_data}
function that uses it to resolve the symbols in string interpolation.

@@

The cases where \eval appears to be overkill include simple expression shapes
such as \k{X}, \k{V}, \k{\$}, \k{<-}, and \k{FUN}. Variable lookup in a remote
environment can be performed by the builtin \k{get} function, a similar case
can be made for assignment and the \k{assign} function. There is a dedicated
\k{do.call} function to perform function calls. However, the unnecessary uses
of \eval are not easy to find and often can only be uncovered by manual
analysis. Dynamic analysis is limited due to coverage issues, and static
analysis of R remains an open problem.

@@

Another example, from \k{coin::copyslots}, constructs an expression that access
fields of an object and copy them.

@@

This can be replaced by symbolic field reads
and writes:

% This should be enough:
\begin{lstlisting}
 slot(target, s) <- slot(source, s)
\end{lstlisting}

% TODO: remove - does not bring anything new or interesting
% As a last example, \k{configr::config.funs.par} constructs three expressions
% that are then fed to \eval. Semantically equivalent code would use \k{get}
% to lookup \k{fun}, and then directly execute the three expressions.

% \begin{lstlisting}
%  config.funs.par <- function(fun="",...) {
%    args.all <- as.list(match.call())
%    args.all <- args.all[names(args.all) != ""]
%    parameters <- list()
%    t1 <- sprintf("parameters <- names(as.list(args(%s)))", fun)
%    t2 <- "parameters <- parameters[parameters != '']"
%    t3 <- "parameters <- args.all[names(args.all) %in% parameters]"
%    for (j in c(t1,t2,t3))
%     eval(parse(text = j))
%    return(parameters)
%  }
% \end{lstlisting}
